Mathematical description of Naomi

### 2022-12-08 src/example_naomi/maths.Rmd
Finish off maths writing (and now naomi.Rmd doesn't render, because I updated the naomi package I think)

* Include two years of ANC testing data: the year of the most recent survey $Y[T_1]$ and the current year $Y[T_2]$
* $W^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients
* $X^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients with ascertained status
* $Y^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients with positive status (either known or tested)
* $Z^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients already on ART prior to first ANC
* Sometimes ANC testing data are only available for part of a given year, for example only the first three quarters. Denote $M^\text{ANC}_{Y[t]} \in \{1, \ldots, 12\}$ the number of months of reported data reflected in counts for year $Y[t]$

@@

* Not sure what $R_x$ means. Is it like $R_x: \mathcal{X} \to \mathcal{X}$
* Some confusion on how to offset terms work -- they seem cool though. Might prefer they not be called $\eta$ which is in my head linear predictor
  * $\eta^\lambda_{R_x, s, a, t}$ aren't these $R_x, s, a, t$ specific rather than 
* Aggregation over age 15-49 uses sum from $a = 15$ to $a = 45$? Perhaps this is because $a = 45$ actually corresponds to the age band 45-49. Edit: yes, $a = 45$ corresponds to 45-49
* Sometimes in subscript we have $x, s, a, t$ and sometimes e.g. $x, s, at$ -- why?
* Why $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ rather than e.g. $\sigma^\lambda \sim \mathcal{N}^+(0, 2.5)$
* Inconsistent use of $T2$ and $T_2$ in the appendix (I prefer the subscript so try to use throughout here)
* In softmax function, confused as to what this equation is saying: shouldn't there be an equals sign in there before the large brackets?

>>>>

* Include two years of ANC testing data: the year of the most recent survey $Y[T_1]$ and the current year $Y[T_2]$
* $W^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients
* $X^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients with ascertained status
* $Y^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients with positive status (either known or tested)
* $Z^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients already on ART prior to first ANC
* Sometimes ANC testing data are only available for part of a given year, for example only the first three quarters. Denote $M^\text{ANC}_{Y[t]} \in \{1, \ldots, 12\}$ the number of months of reported data reflected in counts for year $Y[t]$
* Likelihood for number of ANC clients specified only for year $Y[T_2]$ (**why?**) as
$$
W_{\{x\} Y[T_2]}^{\text{ANC}} \sim \text{Pois} \left( \frac{M_{Y[T 2]}^{\text{ANC}}}{12} \sum_{x \in\{x\}} \sum_{a \in\{15, \ldots 45\}} \Psi_{x, a, T_2} \right)
$$
* Observed number of HIV positive and already on ART among ANC clients at both $Y[T_1]$ and $Y[T_2]$ are modelled by
\begin{aligned}
& Y_{\{x\} Y[t]}^{\text{ANC}} &\sim \text{Bin} \left(X_{\{x\} Y[t]}^{\text{ANC}}, \rho_{\{x\},\{15, \ldots 49\}, t}^{\text{ANC}}\right) \\
& Z_{\{x\} Y[t]}^{\text{ANC}} &\sim \text{Bin} \left(Y_{\{x\} Y[t]}^{\text{ANC}}, \alpha_{\{x\},\{15, \ldots 49\}, t}^{\text{ANC}}\right)
\end{aligned}
where predicted prevalence and ART coverage are aggregated weighted by the predicted number of pregnant women by age $\Psi_{x, a, t}$
\begin{aligned}
\rho_{\{x\}\{a\}, t}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t} \cdot \rho_{x, a, t}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t}} \\
\alpha_{\{x\}\{a\}, t}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t} \cdot \rho_{x, a, t}^{\text{ANC}} \cdot \alpha_{x, a, t}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t} \cdot \rho_{x, a, t}^{\text{ANC}}}
\end{aligned}

## Number receiving ART

* Let $\dot A_{\{x\}, \{s\}, \{a\}, t}$ be data for the number receiving ART **flag: confusion about the notation here. In general would be good to define precisely what is meant by $\{x\}$. Presumably the LHS below is a set, yet the RHS is a sum over the sets so I think just a number -- so how can the two be equal?**
$$
\dot A_{\{x\}, \{s\}, \{a\}, t} = \sum_{s \in \{s\}} \sum_{a \in \{a\}} \sum_{x \in \{x\}} \sum_{x \sim x', x = x'} \dot A_{x', x, s, a, t}
$$
* Model the unobserved numbers of ART clients travelling from $x'$ to $x$ as
$$
\dot A_{x', x, s, a, t} \sim \text{Bin}(N_{x', s, a, t}, \pi_{x', x, s, a, t})
$$
where $\pi_{x', x, s, a, t} = \rho_{x', s, a, t} \cdot \alpha_{x', s, a, t} \cdot \gamma_{x', s, a, t}$ **flag: should this be $\gamma_{x', x, s, a, t}$ since it's people moving from $x'$ to $x$ (and so far no where in the RHS is there an $x$ which is suspicious)**
* This likelihood is approximated using a normal for the sum of binomials
$$
\dot A_{\{x\}, \{s\}, \{a\}, t} \sim \mathcal{N}(\tilde) A_{\{x\}, \{s\}, \{a\}, t}, \sigma^{\tilde A}_{\{x\}, \{s\}, \{a\}, t})
$$
where
  * $\tilde{A}_{\{x\},\{s\}\{a\}, t} = \sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x', x = x'} N_{x', s, a, t} \cdot \pi_{x', x, s, a, t}$
  * $\sigma_{\{x\},\{s\}\{a\}, t}^{\tilde{A}} = \sqrt{\sum_{s \in\{s\}} \sum_{a \in\{a\}} \sum_{x \in\{x\}} \sum_{x \sim x^', x=x \prime} N_{x^', s, a, t} \cdot \pi_{x^', x, s, a, t} \cdot \left(1-\pi_{x^', x, s, a, t}\right)}$

# Identifiability constaints

* Depends on data availability. If data is missing, some parameters might be fixed to default values
* Survey data on ART coverage by age and sex not available then set $u_a^\alpha = 0$ and $u_{a, s = \text{M}}^\alpha = 0$ and use the average age/sex pattern of ART coverage from the Spectrum offset $\eta_{R_x, s, a}^\alpha$
* If there is no ART data at $T_1$ (either survey or ART programme) or no ART data at $T_2$ then change in ART coverage is unidentifiable and set $\beta_{T_2}^\alpha = \beta_{T_2, s = \text{M}}^\alpha = u_{x, {T_2}}^\alpha = u_{x, a < 15, {T_2}}^\alpha = 0$ such that change in ART coverage is modelled by
$$
\text{logit}(\alpha_{x, s, a, {T_2}}) = \text{logit}(\alpha_{x, s, a, {T_1}}) + \eta_{R_x, s, a, {T_2}}^\alpha
$$
* If no ART data (survey or ART programme) are available at $T_1$ or $T_2$ but data on ART coverage among ANC clients are available, the level of ART coverage is not identifiable, but spatial variation is identifiable. In this instance, overall ART coverage is determined by the Spectrum offset, and only area random effects are estimated
\begin{aligned}
\text{logit} \left(\alpha_{x, s, a, T_1} \right) &= u_x^\alpha+\eta_{R_x, s, a}^\alpha \\
\text{logit} \left(\alpha_{x, s, a, T_2} \right) &= \text{logit}\left(\alpha_{x, s, a, T_1}\right)+u_{x, T_2}^\alpha+\eta_{R_{x, s, a, T_2}}^\alpha
\end{aligned}
* If survey data on recent HIV infection are not included in the model, then $\beta_0^\lambda = \beta_S^{\lambda, s = \text{M}} = u_x^\lambda = 0$. The sex ratio for HIV incidence is determined by the sex incidence rate ratio from Spectrum in the same years and the incidence rate in all districts is modelled assuming the same average HIV transmission rate for untreated adults, but varies according to district estimates of HIV prevalence and ART coverage

@@

* Not sure what $R_x$ means. Is it like $R_x: \mathcal{X} \to \mathcal{X}$
* Some confusion on how to offset terms work -- they seem cool though. Might prefer they not be called $\eta$ which is in my head linear predictor
  * $\eta^\lambda_{R_x, s, a, t}$ aren't these $R_x, s, a, t$ specific rather than 
* Aggregation over age 15-49 uses sum from $a = 15$ to $a = 45$? Perhaps this is because $a = 45$ actually corresponds to the age band 45-49. Edit: yes, $a = 45$ corresponds to 45-49
* Sometimes in subscript we have $x, s, a, t$ and sometimes e.g. $x, s, at$ -- why?
* Why $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ rather than e.g. $\sigma^\lambda \sim \mathcal{N}^+(0, 2.5)$
* Inconsistent use of $T2$ and $T_2$ in the appendix (I prefer the subscript so try to use throughout here)
* In softmax function, confused as to what this equation is saying: shouldn't there be an equals sign in there before the large brackets?
* What's $Y[T2][T1]$ is this a typo and should be $Y[T1]$ -- in "The observed number of HIV positive and already on ART..."
* "In cases where some data inputs, some model..." should be "In cases where some data inputs **are missing**, some model..."
